---
title: "lubridate"
output: html_document
date: "2023-06-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
```

The `R` package `lubridate` allows various date manipulations.

## Basic time objects

The most basic one is transforming a date into a `date` object. 

The function `ymd` (year-month-day) is very flexible and can take many different inputs.

Let's format the date **1970-02-01**

```{r}
ymd(19700201)
ymd(700201)
ymd("700201")
ymd("1970-02-01")
```

You can also impute words and in different order (e.g. `dmy` (day month year))

```{r}
dmy("2nd March 1970")
```

The function can be extended to include minutes and secondes

`ymd_hm`

```{r}
ymd_hm("1970-02-01")
```

The function period is more general and allows to flexibly create date objects

```{r}
period(1970, units = c("year"))
period(c(1970, 02), units = c("year", "month"))
period(c(1970, 02, 30), units = c("year", "month", "day"))
period(year = 1970, month = 2)
period(second = 90, minute = 5)
period("2d 2H 2M 2S")
```

## Origin

The date of **origin** with lubridate is `1970-01-01`

```{r}
lubridate::origin
```

The function `as_date` count the number of days since the origin date

```{r}
as_date(0, origin = lubridate::origin)
as_date(1, origin = lubridate::origin)
as_date(11918, origin = lubridate::origin)
```


```{r}
## change ref date
as_date(1, origin = ymd(20000101))
as_date(30, origin = ymd(20000101))
##
```

## Time Zone

All the available time zone in `R` are listed in `OlsonNames()`. 

Find your time zone with

```{r}
Sys.timezone()
# your date and time
Sys.time()
```

You can change time zone by setting the `tz` parameter.

Imagine you have an event at 6pm in Los Angeles, you can find the time of the event in Rome with

```{r}
x <- ymd_hms("2023-06-28 18:00:00", tz = "America/Los_Angeles")
x
with_tz(x, "Europe/Rome")
```

## Adding time

Lubridate allows to add or subtract time very easily. 

Let's start with the date 2000-01-01 and let's add 10 years

```{r}
x <- ymd("2000-01-01")
x + years(10)
```

We can also add days or months easily

```{r}
x + days(365)
```
We can subtract time with

```{r}
x - years(10)
```

Imagine that you start work at 7am and work for 8h30 minutes. 

```{r}
# the work started at #
x <- ymd_hm("2000-06-01-07-00")
x
# lasted 8 hours and 30 mintues
x + hours(8) + minutes(30)
```

## Selecting

You can select the year, month, day, etc. with

```{r}
x <- ymd_hm("2022-06-01-07-00")
year(x)
month(x)
day(x)
```

## Rounding time

Rounding can be useful for calculation, this is done with two functions ``

Imagine you have the date and time: 2022 March 02, 7am 58 minutes and 10 sec

```{r}
x <- ymd_hms("2022 March 02 07am 58min 10")
x
# taking the rounding at the unit below
floor_date(x, unit = "hour")
# taking the rounding at the unit above
ceiling_date(x, unit = "hour")
```


## duration

`Lubridate` uses seconds as the measure for duration. 

Duration functions have a `d` before, like `dseconds`, etc.

```{r}
dseconds(1)
dminutes(1)
dhours(1)
ddays(1)
```
We can transform back to minutes with

(2 hours in minutes)

```{r}
# hours in minutes
dhours(2) / dminutes(1)
```

1 day in minutes is 

```{r}
ddays(1) / dminutes(1)
```

The duration object is the general function for the duration object presented above.

```{r}
# same
dminutes(1)
duration(minutes = 1)
#
duration(minutes = 60)
duration(days = 7)
#
# same
dyears(1)
duration(year = 1)
#
```

